Towards Patient-Specific Neural Field Modeling using Micro-Electrode Data

### 2011-02-11 ltx/data_paper/DataPaper.tex
edits to M step
edits to supp for Xi variables
check updates

Since the state equation is non-linear,  calculation of covariance and  cross-covariance terms required for the maximisation step is not straightforward as in the linear case. Therefore, an additional  Taylor expansion step to the URTSS algorithm is also created to approximate $q(\mathbf{x}_t)$ at each time instant which is to be used in the M-step. This is accomplished by taking the first-order truncated Taylor-series expansion of the sigmoidal activation function  around the smoothed state, $\hat{\mathbf x}_t^{b}$ at each time instant, i.e.
\begin{equation}\label{eq:SigmoidTaylor}
 f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\mathbf x_t\right) \approx f\left(\boldsymbol \phi^\top(\mathbf r')\hat{\mathbf x}_t^{b}\right)+  \boldsymbol \phi^\top\left(\mathbf r'\right)\left(\mathbf x_t - \hat{\mathbf x}_t^{b}\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right)
\end{equation}
where 
\begin{equation}\label{eq:SigmoidDerivative}
 f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right)=\varsigma f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right)\left( 1-f\left( \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right)\right)
\end{equation}
substituting \ref{eq:SigmoidTaylor} in \ref{eq:QmatrixForSigmapoints}  we get an approximation for $q(\mathbf x_t)$ around the smoothed estimate $\hat{\mathbf{x}}_t^b$ 
\begin{equation}\label{eq:qTaylor}
 q(\mathbf x_t) \approx q\left(\hat{\mathbf x}_t^{b}\right)+\int_\Omega \boldsymbol{\Psi}\left(\mathbf{r}'\right)\boldsymbol \phi^\top\left(\mathbf r'\right) \left(\mathbf x_t -\hat{\mathbf x}_t^{b}\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right) d\mathbf{r}'
\end{equation}
Equation \ref{eq:qTaylor} is used to approximate the lower bound on the log-likelihood function in the M-step.

@@

 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&\approx& -(T-1)n_y\ln \sigma_{\epsilon}^2-\mathrm{tr}\left\lbrace\boldsymbol\Sigma_{\epsilon}^{-1}\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-&(T-1)n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \Xi'_0 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +2\frac{1}{\sigma_d^2}\Xi_0\boldsymbol\theta+
\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace \nonumber \\
&-&\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \Xi_2\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \Xi_3 \boldsymbol\theta -  \frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\Xi_4\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace 

@@

	\Xi'_0&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf P_{t+1}+\mathbf{\hat x}_{t+1}\mathbf{\hat x}_{t+1}^\top\right)\label{eq:def of Xi'0} \\
\Xi_0&=\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_{t+1} \tilde{\boldsymbol\Sigma}_e^{-1} \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] \\	
\Xi_1&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\
\Xi_2&=\sum_{t=0}^{T-1}\left[q^\top(\mathbf{\hat x}_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\Lambda_t(\mathbf r)\mathbf P_t \tilde{\Lambda}_t(\mathbf r)\right] \\
 \Xi_3&=\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf P_t \tilde{\boldsymbol\Sigma}_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] \\
 \Xi_4&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right)

@@

	 \Lambda_t(\mathbf r)&=\begin{bmatrix}\boldsymbol \lambda_t^{(1)}(\mathbf r) & \boldsymbol \lambda_t^{(2)}(\mathbf r)& \dots &\boldsymbol \lambda_t^{(i)}(\mathbf r)\end{bmatrix} \\
	 \boldsymbol\lambda_t^{(i)}(\mathbf r)&=\int_{\Omega} \boldsymbol \Psi^\top(\mathbf r')\tilde{\boldsymbol\Sigma}_e^{-1}\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r'\\
	\tilde{\Lambda}_t(\mathbf r)&=\begin{bmatrix}\tilde{\boldsymbol \lambda}_t^{(1)}(\mathbf r) \\ \tilde{\boldsymbol \lambda}_t^{(2)}(\mathbf r) \\ \vdots \\ \tilde{\boldsymbol \lambda}_t^{(i)}(\mathbf r)\end{bmatrix} \\
	 \tilde{\boldsymbol \lambda}_t^{(i)}(\mathbf r)&=\int_{\Omega} \boldsymbol \Psi(\mathbf r')\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r'

@@

It should be noted that the second term in equation \ref{eq:Voldermont} is obtained by adding and subtracting $\mathbf C \mathbf{\hat{x}}_t \mathbf{\hat{x}}_t^\top \mathbf C^\top$ to the expectation, $\mathbf E_{\boldsymbol\Theta'}\left[ .\right] $. Given the state estimates, $\mathbf {\hat x^{b}}_t$ the state covariance $\mathbf P_t^b$ and cross-covariance $\mathbf M_t^b$ , the $\Xi$-variables can be computed directly.

When taking the derivative of equation~\ref{eq:Voldermont}, the constant $\beta$ described in equation~\ref{eq:def of beta} will disappear. Differntiating with respect to $\boldsymbol\theta$ we have
\begin{equation}\label{eq:Qpartialtheta}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\theta}=2\Xi_0-\boldsymbol\theta^\top(\Xi_2^\top+\Xi_2)-2\xi\Xi_3
\end{equation}
noting that $\Xi_2$ is symmetric and equating \ref{eq:Qpartialtheta} to the zero vector we get
\begin{equation}\label{eq:theta}
 \boldsymbol \theta=\Xi_2^{-1}\left(\Xi_0^\top-\xi\Xi_3^\top \right)
\end{equation}
similarly differentiating with respect to $\xi$ we have
\begin{equation}\label{eq:Qpartialxi}
 \frac{\partial \mathcal Q}{\partial \xi}=2tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace -2 \Xi_3 \boldsymbol\theta - 2\xi tr \left\lbrace\Xi_4\boldsymbol\Sigma_e^{-1} \right\rbrace
\end{equation}
substituting \ref{eq:theta} in \ref{eq:Qpartialxi} and equating to zero we get
\begin{equation}
 \xi=\frac{tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_0^\top}{tr\left\lbrace \Xi_4 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_3^\top}
\end{equation}
Rewrite equation~\ref{eq:Voldermont} as
\begin{equation}
 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')\propto-(T-1)n_y\ln\sigma^2_{\epsilon}-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace
\end{equation}
differentiating with respect to $\sigma_{\epsilon}^2$
\begin{equation}
  \frac{\partial \mathcal Q}{\partial \sigma_{\epsilon}^2}=-(T-1)n_y\frac{1}{\sigma_{\epsilon}^2}+\frac{1}{\sigma_{\epsilon^4}}\mathrm{tr}\left\lbrace\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace
\end{equation}
setting above to zero and solving for $\sigma_{\epsilon}^2$ we have
\begin{equation}
 \sigma_{\epsilon}^2=\frac{1}{(T-1)n_y}\sum_{t=0}^{T-1}\mathrm{tr}\left\lbrace (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top \right\rbrace
\end{equation}

>>>>

Since the state equation is non-linear,  calculation of covariance and  cross-covariance terms required for the maximisation step is not straightforward as in the linear case. Therefore, an additional  Taylor expansion step to the URTSS algorithm is also created to approximate $q(\mathbf{x}_t)$ at each time instant which is to be used in the M-step. This is accomplished by taking the first-order truncated Taylor-series expansion of the sigmoidal activation function  around the smoothed state, $\hat{\mathbf x}_t^{b}$ at each time instant, i.e.
\begin{equation}\label{eq:SigmoidTaylor}
 f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\mathbf x_t\right) \approx f\left(\boldsymbol \phi^\top(\mathbf r')\hat{\mathbf x}_t\right)+  \boldsymbol \phi^\top\left(\mathbf r'\right)\left(\mathbf x_t - \hat{\mathbf x}_t\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)
\end{equation}
where 
\begin{equation}\label{eq:SigmoidDerivative}
 f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)=\varsigma f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)\left( 1-f\left( \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)\right)
\end{equation}
substituting \ref{eq:SigmoidTaylor} in \ref{eq:QmatrixForSigmapoints}  we get an approximation for $q(\mathbf x_t)$ around the smoothed estimate $\hat{\mathbf{x}}_t^b$ 
\begin{equation}\label{eq:qTaylor}
 q(\mathbf x_t) \approx q\left(\hat{\mathbf x}_t\right)+\int_\Omega \boldsymbol{\Psi}\left(\mathbf{r}'\right)\boldsymbol \phi^\top\left(\mathbf r'\right) \left(\mathbf x_t -\hat{\mathbf x}_t\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right) d\mathbf{r}'
\end{equation}
Equation \ref{eq:qTaylor} is used to approximate the lower bound on the log-likelihood function in the M-step.

@@

 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&\approx& -(T-1)n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-&(T-1)n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \Xi_{0} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \Xi_{1} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace+2\frac{1}{\sigma_d^2}\Xi_{2}\boldsymbol\theta
 \nonumber \\
&-&\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \Xi_{3}\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \Xi_{4} \boldsymbol\theta -  \frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\Xi_{5}\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace 

@@

	\Xi_{0}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf P_{t+1}+\mathbf{\hat x}_{t+1}\mathbf{\hat x}_{t+1}^\top\right)\label{eq:defofXi0} \\
\Xi_{1}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \label{eq:defofXi1} \\
\Xi_{2}&=\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_{t+1} \tilde{\boldsymbol\Sigma}_e^{-1} \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] \\	
\Xi_{3}&=\sum_{t=0}^{T-1}\left[q^\top(\mathbf{\hat x}_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\Lambda_t(\mathbf r)\mathbf P_t \tilde{\Lambda}_t(\mathbf r)\right] \\
 \Xi_{4}&=\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf P_t \tilde{\boldsymbol\Sigma}_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] \\
 \Xi_{5}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right)

@@

	 \Lambda_t(\mathbf r)&=\begin{bmatrix}\boldsymbol \lambda_t^{(1)}(\mathbf r) & \boldsymbol \lambda_t^{(2)}(\mathbf r)& \dots &\boldsymbol \lambda_t^{(i)}(\mathbf r)\end{bmatrix} \label{eq:Lambda}\\
	 \boldsymbol\lambda_t^{(i)}(\mathbf r)&=\int_{\Omega} \boldsymbol \Psi^\top(\mathbf r')\tilde{\boldsymbol\Sigma}_e^{-1}\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r'\label{eq:lambdai}\\
	\tilde{\Lambda}_t(\mathbf r)&=\begin{bmatrix}\tilde{\boldsymbol \lambda}_t^{(1)}(\mathbf r) \\ \tilde{\boldsymbol \lambda}_t^{(2)}(\mathbf r) \\ \vdots \\ \tilde{\boldsymbol \lambda}_t^{(i)}(\mathbf r)\end{bmatrix} \label{eq:Lambdatilde}\\
	 \tilde{\boldsymbol \lambda}_t^{(i)}(\mathbf r)&=\int_{\Omega} \boldsymbol \Psi(\mathbf r')\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r' \label{eq:lambdatildei}

@@

It should be noted that the second term in equation \ref{eq:Voldermont} is obtained by adding and subtracting $\mathbf C \mathbf{\hat{x}}_{t+1} \mathbf{\hat{x}}_{t+1}^\top \mathbf C^\top$ to the expectation, $\mathbf E_{\boldsymbol\Theta'}\left[ .\right] $. Given the state estimates, $\mathbf {\hat x^{b}}_t$ the state covariance $\mathbf P_t^b$ and cross-covariance $\mathbf M_t^b$ , the $\Xi$-variables can be computed directly. In the M-step, an update $\boldsymbol\Theta'$ is chosen such that $\mathcal{Q}\left(\boldsymbol\Theta',\boldsymbol\Theta'\right)>\mathcal{Q}\left(\boldsymbol\Theta,\boldsymbol\Theta'\right)$, this will guarantee that the likelihood increases.

In order to find the maximum of the $\mathcal{Q}$-function Differntiating with respect to $\boldsymbol\theta$ we have
\begin{equation}\label{eq:Qpartialtheta}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\theta}=2\Xi_0-\boldsymbol\theta^\top(\Xi_2^\top+\Xi_2)-2\xi\Xi_3
\end{equation}
noting that $\Xi_2$ is symmetric and equating \ref{eq:Qpartialtheta} to the zero vector we get
\begin{equation}\label{eq:theta}
 \boldsymbol \theta=\Xi_2^{-1}\left(\Xi_0^\top-\xi\Xi_3^\top \right)
\end{equation}
similarly differentiating with respect to $\xi$ we have
\begin{equation}\label{eq:Qpartialxi}
 \frac{\partial \mathcal Q}{\partial \xi}=2tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace -2 \Xi_3 \boldsymbol\theta - 2\xi tr \left\lbrace\Xi_4\boldsymbol\Sigma_e^{-1} \right\rbrace
\end{equation}
substituting \ref{eq:theta} in \ref{eq:Qpartialxi} and equating to zero we get
\begin{equation}
 \xi=\frac{tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_0^\top}{tr\left\lbrace \Xi_4 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_3^\top}
\end{equation}
Rewrite equation~\ref{eq:Voldermont} as
\begin{equation}
 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')\propto-(T-1)n_y\ln\sigma^2_{\epsilon}-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace
\end{equation}
differentiating with respect to $\sigma_{\epsilon}^2$
\begin{equation}
  \frac{\partial \mathcal Q}{\partial \sigma_{\epsilon}^2}=-(T-1)n_y\frac{1}{\sigma_{\epsilon}^2}+\frac{1}{\sigma_{\epsilon^4}}\mathrm{tr}\left\lbrace\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace
\end{equation}
setting above to zero and solving for $\sigma_{\epsilon}^2$ we have
\begin{equation}
 \sigma_{\epsilon}^2=\frac{1}{(T-1)n_y}\sum_{t=0}^{T-1}\mathrm{tr}\left\lbrace (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top \right\rbrace
\end{equation}
